POPLMark goes Concurrent!\\Challenge problems for process calculi and behavioural types

### 2024-02-23 challenges/linearity.tex
Further edits to draft

\begin{lemma}[Weakening]\mbox{}
  \label{lemma:weak}
  \begin{enumerate}
  \item If \( \types{\Gamma; \Delta}{P} \) then
    \( \types{(\Cadd{\Gamma}{l});\Delta}{P} \).
  \item If \( \types{\Gamma; \Delta}{P} \) % and \( \tend{\Delta} \)
    then
    \( \types{\Gamma;(\Cadd{\Delta}{\hastype{x}{\Tend}})}{P} \).
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on the given derivations.
\end{proof}
\begin{lemma}[Strengthening]\mbox{}
  \label{lemma:strenD}
  If \( \types{\Gamma; (\Cadd{\Delta}{\hastype{x}{T}})}{P} \) and
  $x\not\in \freenames{P}$, then \( \types{\Gamma; \Delta}{P} \).
\end{lemma}
\begin{proof}
  By induction on the derivation of
%%  last applied rule of the proof of --AM changed for uniformity
  $\types{\Gamma; (\Cadd{\Delta}{\hastype{x}{T}})}{P}$.
  % , for each case inverting $\types{\Gamma; (\Delta, x : T)}{P}$ to
  % the uniquely applicable rule.
  \begin{description}
  \item[\TirName{T-End}] Since $\tend{\Delta}$, we can just reapply
    the rule without $\hastype{x}{T}$.
    % [$P = \Pend$] we Apply \TirName{T-Inact}.

@@

\begin{lemma}[Preservation for $\stackrel{a}{\equiv}$]
  \label{le:presequiva}
  If \( \sconga{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By case analysis on the \TirName{Sc} rule applied:
  \begin{description}
  \item[\TirName {Par-Comm/Assoc}] By rearranging sub-derivations noting that
    order does not matter for linear contexts
  \item[\TirName{Par-Inact}]
   Right-to-left by strengthening~\ref{lemma:strenD}. Vice-versa,
    by picking $T$ to be $\Tend$ and weakening~\ref{lemma:weak} part 2.
  \item[\TirName{Res-Par}] By case analysis on $x : T$ being linear or $\Tend$ and applying weakening and strengthening accordingly.
  \item[\TirName{Res-Inact:}] Use weakening %~\ref{lemma:weak} part 2
  \item[\TirName{Res}] Noting that order does not matter
  \end{description}
\end{proof}

@@

\begin{lemma}[Preservation for $\equiv$]
  \label{le:presequiv}
    If \( \scong{P}{Q} \) then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By induction on the structure of the derivation of \( \scong{P}{Q} \), with an inner induction of the structure of a process context.
  \begin{description}
  \item[\TirName {Refl}] Immediate
      \item[\TirName {Sym}] By IH.
      \item[\TirName {Trans}]  By two appeal to the IH
      \item[\TirName {Cong}] By induction on the structure of $C$. If
        the context is a hole, apply Lemma~\ref{le:presequiva}. In the
        step case, apply the IH: for example if $C$ has the form
        ${\Ppar {C'} R}$, noting that ${\PBpar{ C'} R}[P]$ is equal to
        ${\Ppar {C'[P]} R}$ we have
        \( \types{\Gamma;\Delta}{{\PBpar {C'} R}[P]} \) iff
        \( \types{\Gamma;\Delta}{{\PBpar {C'} R}[Q]} \) by rule T-Par and
        the IH.
  \end{description}
\end{proof}

\begin{theorem}[Subject reduction]
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \) then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reduces{P}{Q}$. The cases \TirName{R-Par} and \TirName{R-Res} follow immediately by IH. Case \TirName{R-Struct} appeals twice to
  preservation of $\equiv$ (Lemma~\ref{le:presequiv}) and to the
  IH. For \TirName{R-Com}, suppose that \TirName{T-Res} introduces in $\Delta$ the
  assumptions $\Cadd{\hastype{x}{\Tout{U}}}{\hastype{y}{\Tin{\dual{U}}}}$.  Building the only derivation for the
  hypothesis, we know that $\Delta = \Delta_1, \Delta_2, \Delta_3$
  where $\types{\Gamma;\Delta_3}{R}$. We also have
  $\types{\Gamma;(\Cadd{\Delta_1}{\hastype{x}{U}})}{P}$, ${\cal D}_{2} $ a proof of
  $\types{\Gamma,l;(\Cadd{\Delta_2}{\hastype{y}{\dual{U}}})}{Q}$ and $\cal V $ a proof of
  $\typev{\Gamma}{\hastype{a}{\Tbase}}$.  From ${\cal D}_{2}$ and $\cal V$ we use the
  substitution \cref{le:subst} to obtain $\types{\Gamma;\Cadd{\Delta_2}{\hastype{y}{\dual{U}}}}{\subst{Q}{a}{l}}$. We then conclude the proof with rules \TirName{T-Par} (twice) and \TirName{T-Res}.
\end{proof}

@@

\begin{corollary}
  If \( \types{\Gamma;\Cempty}{P} \) and \( \reduces{P}{Q} \) then \( Q \) is well formed.
\end{corollary}

>>>>

\begin{lemma}[Weakening]\mbox{}
  \label{lemma:weak}
  \begin{enumerate}
  \item If \( \types{\Gamma; \Delta}{P} \), then
    \( \types{(\Cadd{\Gamma}{l});\Delta}{P} \).
  \item If \( \types{\Gamma; \Delta}{P} \), % and \( \tend{\Delta} \)
    then
    \( \types{\Gamma;(\Cadd{\Delta}{\hastype{x}{\Tend}})}{P} \).
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on the given derivations.
\end{proof}
\begin{lemma}[Strengthening]\mbox{}
  \label{lemma:strenD}
  If \(\types{\Gamma; (\Cadd{\Delta}{\hastype{x}{T}})}{P}\) and
  $x\not\in \freenames{P}$, then \(\types{\Gamma; \Delta}{P}\).
\end{lemma}
\begin{proof}
  By induction on the derivation of
%%  last applied rule of the proof of --AM changed for uniformity
  $\types{\Gamma; (\Cadd{\Delta}{\hastype{x}{T}})}{P}$.
  % , for each case inverting $\types{\Gamma; (\Delta, x : T)}{P}$ to
  % the uniquely applicable rule.
  \begin{description}
  \item[\TirName{T-End}] Since $\tend{\Delta}$, we can just reapply
    the rule without $\hastype{x}{T}$.
    % [$P = \Pend$] we Apply \TirName{T-Inact}.

@@

\begin{lemma}[Preservation for $\stackrel{a}{\equiv}$]
  \label{le:presequiva}
  If \( \sconga{P}{Q} \), then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By case analysis on the \TirName{Sc} rule applied:
  \begin{description}
  \item[\TirName {Par-Comm/Assoc}] By rearranging sub-derivations noting that
    order does not matter for linear contexts
  \item[\TirName{Par-Inact}]
   Right-to-left by strengthening~\ref{lemma:strenD}. Vice-versa,
    by picking $T$ to be $\Tend$ and weakening~\ref{lemma:weak} part 2.
  \item[\TirName{Res-Par}] By case analysis on $x : T$ being linear or $\Tend$ and applying weakening and strengthening accordingly.
  \item[\TirName{Res-Inact:}] Use weakening %~\ref{lemma:weak} part 2
  \item[\TirName{Res}] Noting that order does not matter
  \end{description}
\end{proof}

@@

\begin{lemma}[Preservation for $\equiv$]
  \label{le:presequiv}
    If \( \scong{P}{Q} \), then \( \types{\Gamma;\Delta}{P} \)  iff \( \types{\Gamma;\Delta}{Q} \).
\end{lemma}
\begin{proof}
  By induction on the structure of the derivation of \( \scong{P}{Q} \), with an inner induction of the structure of a process context.
  \begin{description}
  \item[\TirName {Refl}] Immediate
      \item[\TirName {Sym}] By IH.
      \item[\TirName {Trans}]  By two appeal to the IH
      \item[\TirName {Cong}] By induction on the structure of $C$. If
        the context is a hole, apply Lemma~\ref{le:presequiva}. In the
        step case, apply the IH: for example if $C$ has the form
        ${\Ppar {C'} R}$, noting that ${\PBpar{ C'} R}[P]$ is equal to
        ${\Ppar {C'[P]} R}$ we have
        \( \types{\Gamma;\Delta}{{\PBpar {C'} R}[P]} \) iff
        \( \types{\Gamma;\Delta}{{\PBpar {C'} R}[Q]} \) by rule T-Par and
        the IH.
  \end{description}
\end{proof}

\begin{theorem}[Subject reduction]
  If \( \types{\Gamma;\Delta}{P} \) and \( \reduces{P}{Q} \), then \( \types{\Gamma;\Delta}{Q} \).
\end{theorem}
\begin{proof}
  By induction on the derivation of $\reduces{P}{Q}$. The cases \TirName{R-Par} and \TirName{R-Res} follow immediately by IH. Case \TirName{R-Struct} appeals twice to
  preservation of $\equiv$ (Lemma~\ref{le:presequiv}) and to the
  IH. For \TirName{R-Com}, suppose that \TirName{T-Res} introduces in $\Delta$ the
  assumptions $\Cadd{\hastype{x}{\Tout{U}}}{\hastype{y}{\Tin{\dual{U}}}}$.  Building the only derivation for the
  hypothesis, we know that $\Delta = \Delta_1, \Delta_2, \Delta_3$
  where $\types{\Gamma;\Delta_3}{R}$. We also have
  $\types{\Gamma;(\Cadd{\Delta_1}{\hastype{x}{U}})}{P}$, ${\cal D}_{2} $ a proof of
  $\types{\Gamma,l;(\Cadd{\Delta_2}{\hastype{y}{\dual{U}}})}{Q}$ and $\cal V $ a proof of
  $\typev{\Gamma}{\hastype{a}{\Tbase}}$.  From ${\cal D}_{2}$ and $\cal V$ we use the
  substitution \cref{le:subst} to obtain $\types{\Gamma;\Cadd{\Delta_2}{\hastype{y}{\dual{U}}}}{\subst{Q}{a}{l}}$. We then conclude the proof with rules \TirName{T-Par} (twice) and \TirName{T-Res}.
\end{proof}

@@

\begin{corollary}
  If \( \types{\Gamma;\Cempty}{P} \) and \( \reduces{P}{Q} \), then \( Q \) is well formed.
\end{corollary}

### 2024-02-23 challenges/scope-extrusion.tex
Edits and comments
Further edits to draft

\begin{lemma}\label{se-lemma-normalized-reduction}
  If \( \reduces{P}{Q} \) then there are \( x, y, z, z_1, \dots, z_n, R_1, R_2, \) and \( S\) such that
  \begin{align*}
    &\scong{P}{\Pres{z_1}{\!\dots \Pres{z_n}{(\Ppar{(\Ppar{\Pout{x}{y}{R_1}}{\Pin{x}{z}{R_2}})}{S})}}} \\
    &\scong{Q}{\Pres{z_1}{\!\dots \Pres{z_n}{(\Ppar{(\Ppar{R_1}{\subst{R_2}{y}{z}})}{S})}}}
  \end{align*}
\end{lemma}
\begin{proof}
  Follows immediately from \cref{se-lemma-normalized-derivation} and the shape of a normalized derivation.
\end{proof}

The objective of this challenge is to prove the following theorems, which together show the equivalence between the reduction semantics and the transition system semantics up to structural congruence.
The first of the theorems involves reasoning about scope extrusion more directly than the other, and if time does not permit proving both of the theorems, \cref{thm:se-trans-implies-red} should be proven first.
\begin{theorem}\label{thm:se-trans-implies-red}
  \( \transition{P}{\Atau}{Q} \) implies \( \reduces{P}{Q} \).
\end{theorem}
\begin{proof}[Sketch]
  The proof is by induction on the inference of \( \transition{P}{\Atau}{Q} \) using the following lemmas:
  \begin{enumerate}
  \item if \( \transition{Q}{\Aoutf{x}{y}}{Q'} \) then \( \scong{Q}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\Pout{x}{y}{R}}{S})}}} \) and \( \scong{Q'}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{R}{S})}}} \) where \( x,y \notin \{ w_1, \dots, w_n \} \).
  \item if \( \transition{Q}{\Aoutb{x}{z}}{Q'} \) then \( \scong{Q}{\Pres{z}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\Pout{x}{z}{R}}{S})}}}} \) and \( \scong{Q'}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{R}{S})}}} \) where \( x \notin \{ z, w_1, \dots, w_n \} \).
  \item if \( \transition{Q}{\Ain{x}{y}}{Q'} \) then \( \scong{Q}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\Pin{x}{z}{R}}{S})}}} \) and \( \scong{Q'}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\subst{R}{y}{z}}{S})}}} \) where \( x \notin \{ w_1, \dots, w_n \} \).
  \end{enumerate}
\end{proof}

>>>>

\begin{lemma}\label{se-lemma-normalized-reduction}
  If \( \reduces{P}{Q} \), then there are \( x, y, z, z_1, \dots, z_n, R_1, R_2, \) and \( S\) such that
  \begin{align*}
    &\scong{P}{\Pres{z_1}{\!\dots \Pres{z_n}{(\Ppar{(\Ppar{\Pout{x}{y}{R_1}}{\Pin{x}{z}{R_2}})}{S})}}} \\
    &\scong{Q}{\Pres{z_1}{\!\dots \Pres{z_n}{(\Ppar{(\Ppar{R_1}{\subst{R_2}{y}{z}})}{S})}}}
  \end{align*}
\end{lemma}
\begin{proof}
  Follows immediately from \cref{se-lemma-normalized-derivation} and the shape of a normalized derivation.
\end{proof}

The objective of this challenge is to prove the following theorems, which together show the equivalence between the reduction semantics and the transition system semantics up to structural congruence.
The first of the theorems involves reasoning about scope extrusion more directly than the other, and if time does not permit proving both of the theorems, \cref{thm:se-trans-implies-red} should be proven first.
\begin{theorem}\label{thm:se-trans-implies-red}
  \( \transition{P}{\Atau}{Q} \) implies \( \reduces{P}{Q} \).
\end{theorem}
\begin{proof}[Sketch]
  The proof is by induction on the inference of \( \transition{P}{\Atau}{Q} \) using the following lemmata:
  \begin{enumerate}
  \item if \( \transition{Q}{\Aoutf{x}{y}}{Q'} \) then \( \scong{Q}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\Pout{x}{y}{R}}{S})}}} \) and \( \scong{Q'}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{R}{S})}}} \) where \( x,y \notin \{ w_1, \dots, w_n \} \).
  \item if \( \transition{Q}{\Aoutb{x}{z}}{Q'} \) then \( \scong{Q}{\Pres{z}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\Pout{x}{z}{R}}{S})}}}} \) and \( \scong{Q'}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{R}{S})}}} \) where \( x \notin \{ z, w_1, \dots, w_n \} \).
  \item if \( \transition{Q}{\Ain{x}{y}}{Q'} \) then \( \scong{Q}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\Pin{x}{z}{R}}{S})}}} \) and \( \scong{Q'}{\Pres{w_1}{\!\dots \Pres{w_n}{(\Ppar{\subst{R}{y}{z}}{S})}}} \) where \( x \notin \{ w_1, \dots, w_n \} \).
  \end{enumerate}
\end{proof}
